---
title: "EDA of children in the United States"
author: "WeiPan Cai, Yue Wang, Tian Sheng, Jiwen You, Sen Zhuang"
output:
  html_notebook: default
  html_document: default
---

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that we end up using
# in your analysis in this chunk
library(data.table)
library(ggplot2)
library(sqldf)
library(maps)
library(plyr)
library(googleVis)
library(dplyr)
library(plotly)
```

#Summary
In this project, we are interested in what factors will affect the number of children. So based on the household data, we choose varibles like State, family type and employment status, languages spoken in the family, household income. In addtion, we combined several variables like water fees to represent living costs.

#Inital data overview
```{r}
# Load RData file
load(file.choose())

# Change variables to their correct class
# df$NOC <- as.factor(df$NOC) # Convert number of own children to factor
# df$NP <- as.factor(df$NP)
# df$BDSP <- as.factor(df$BDSP)
# df$RMSP <- as.factor(df$RMSP)
# df$FES <- as.factor(df$FES)
# df$HHL <- as.factor(df$HHL)
# df$WKEXREL <- as.factor(df$WKEXREL)
# df$WORKSTAT <- as.factor(df$WORKSTAT)

# print out summary stats
str(df)

```
Our dataset consists of 19 variables and total of 308611 households.

In the first part, we will take a look at the NOC in different states.
# Part One (Sen Zhuang)
```{r}

```

From the first part, we have seen that "Utah","Idaho","Alaska","North Dakota" and "South Dakota" have higher number of children. "Rhode Island","Vermont","West Virginia","New hampshire" and "District of Columbia" have lower number of children. Hence, in the folloing part, we want to find what factors will affect the NOC. We are interested in Languages spoken in the family, Family Type and Employment Status(FES), Household Income and living costs. We start by studying the relationship between NOC and each individaul factor. Then, focusing on the first five states which have higher number of children and the last five states which have lower number of children, we study our factors further in these states and try to find some paterns.

# Part Two
a) Languages spoken in family (Weipan Cai)
```{r}

```

b) Family Type and Employment Status (Yue Wang)

```{r, echo=FALSE}
# study the relationship between FES and NOC
working <- select(df, NOC, FES)
work1 <- working %>%
          na.omit()
size <- count(work1, FES)
# compare mean of different FES
result <- aggregate(NOC~FES, data = work1, FUN=mean)
# FamilyTypeEmploymentStatus code
fescode <- "FES,FamilyTypeEmploymentStatus
1, Both in Labor Force
2, Only Husband in Labor Force
3, Only Wife in Labor Force
4, Neither in Labor Force
5, Male in Labor Force
6, Male not in Labor Force
7, Female in Labor Force
8, Female not in Labor Force
"
codeoffes <- fread(fescode)
```

The bubble chart below shows the mean NOC under different family type and employment status. From it, we can find that Married-couple, only husband in Labor Force that has a biggest mean number of childern.
```{r}
# bubble chart
d <- diamonds[sample(nrow(diamonds),1000),]
plot_ly(d,x=result$FES, y=result$NOC, size=size$n, mode="markers") %>%
  layout(title = "Bubble chart of FES and mean of NOC", xaxis = list(title = "FES"), yaxis = list(title = "NOC"))
```

In the following part, I use ggplot to show the percentage and total NOC under different family type and employment status. From these plots, we can find that the family type that "Only Husband in Labor Force" tend to have more number of children. The family type that "Female in Labor Force" tend to have less number of children.
```{r}
work1 = tbl_df(work1)
# work2 is the table group_by(NOC,FES)
work2 <- work1 %>% group_by(NOC,FES) %>% summarise(n=n()) %>% mutate(percentage = n/sum(n) * 100)
work2 <- mutate(work2, percentage = round(percentage,2))
work2 <- sqldf('SELECT work2.NOC, work2.FES,work2.n,work2.percentage, codeoffes.FamilyTypeEmploymentStatus from work2,codeoffes where work2.FES=codeoffes.FES')
# by percentage
ggplot(work2,aes(x=NOC,y=percentage,fill=factor(FamilyTypeEmploymentStatus)))+xlab("Number of children")+ylab("Percentage")+geom_bar(stat="identity")+ggtitle("Percentage of FES under different NOC")
# by number
ggplot(work2,aes(x=NOC,y=n,fill=factor(FamilyTypeEmploymentStatus)))+xlab("Number of children")+ylab("Total Number")+geom_bar(stat="identity")+ggtitle("Total Number of FES under different NOC")
# work3 is the table group_by(FES,NOC)
work3 <- work1 %>% group_by(FES,NOC) %>% summarise(n=n()) %>% mutate(percentage = n/sum(n) * 100)
work3 <- mutate(work3,percentage = round(percentage,2))
work3 <- sqldf('SELECT work3.FES, work3.NOC,work3.n,work3.percentage, codeoffes.FamilyTypeEmploymentStatus from work3,codeoffes where work3.FES=codeoffes.FES')
# by percentage 
ggplot(work3,aes(x=FES,y=percentage,fill=factor(NOC)))+xlab("Family Type and Employment Status ")+ylab("Percentage")+geom_bar(stat="identity")+ggtitle("Percentage of NOC under different FES") + theme(legend.position="bottom")
# by number
ggplot(work3,aes(x=FES,y=n,fill=factor(NOC)))+xlab("Family Type and Employment Status ")+ylab("Total number")+geom_bar(stat="identity")+ggtitle("Total number of NOC under different FES") + theme(legend.position="bottom")
```

Then I use Sankey plot to display the relationship between Family Type Employment Status and number of children. In addition, from the ggplot above, we can find that few family will have more than four children, so our focus will be on no more than four children and married couple family type.
```{r}
# sankey plot
SKdata <- data.frame(From=c(rep("Both in Labor Force",12),rep("Only Husband in Labor Force",12),rep("Only Wife in Labor Force",12),rep("Neither in Labor Force",12),
                            rep("Male in Labor Force",12),rep("Male not in Labor Force",12),rep("Female in Labor Force",12),rep("Female not in Labor Force",12)),
                     To=c(rep(c("1","2","3","4","5","6","7","8","9","10","11","12"),8)),
                     Weight=c(40.92,41.22,13.72,3.17,0.68,0.18,0.07,0.03,0.01,0,0,0,31.47,38.59,19.52,7.11,1.99,0.77,0.29, 0.15,0.07,0.03,0.01,0,47.26,35.25,12.27,3.57,1.05,0.33,
                              0.18,0.06,0.02,0.01,0,0,49.46,30.02,12.32,5.53,1.34,0.72,0.41,0.08,0.08,0.03,0.03,0,55.22,30.64,10.24,2.93,0.74,0.16,0.06,0.01,0,0,0,0,62.77,24.31,8.81,
                              2.96,0.76,0.25,0.13,0,0,0,0,0,51.55,31.85,11.90,3.45,0.89,0.26,0.07,0.02,0.01,0,0,0,43.74,31.90,15.28,6.07,1.97,0.73,0.19,0.09,0.02,0.02,0,0.01))
Sankey <- gvisSankey(SKdata, from="From", to="To",weight="Weight",options=list(sankey="{link: {color: { fill: '#d799ae' } },
                            node: { color: { fill: '#a61d4c' },
                            label: { color: '#871b47' } }}"))
plot(Sankey)

# sankey plot about no more than four children and married couple family type
SKdata1 <-data.frame(From=c(rep("Both in Labor Force",4),rep("Only Husband in Labor Force",4),rep("Only Wife in Labor Force",4),rep("Neither in Labor Force",4)),
                     To=c(rep(c("1","2","3","4"),4)),
                     Weight=c(40.92,41.22,13.72,3.17,31.47,38.59,19.52,7.11,47.26,35.25,12.27,3.57,49.46,30.02,12.32,5.53))
Sankey1 <- gvisSankey(SKdata1, from="From", to="To",weight="Weight",options=list(sankey="{link: {color: { fill: '#d799ae' } },
                            node: { color: { fill: '#a61d4c' },
                            label: { color: '#871b47' } }}"))
plot(Sankey1)
```

Based on the first part, I try to compare the Family Type and Employment Status in states which have more children and states which have less children. 
```{r}
# Then we want to dig out more about the FES in different status.
high <- sqldf('SELECT df.NOC, df.FES, df.name from df where name in ("Utah","Idaho","Alaska","North Dakota","South Dakota")')
low <- sqldf('SELECT NOC, FES, name from df where name in ("District of Columbia","New Hampshire","West Virginia","Vermont","Rhode Island")')
# Then we want to study the relationship between state and FES
# study the fist five rank high states
high1 <- select(high, FES,name) %>% na.omit()
statehigh <- high1 %>% group_by(name,FES) %>% summarise(n=n()) %>% mutate(percentage = n/sum(n) * 100)
statehigh <- mutate(statehigh,percentage = round(percentage,2))
statehigh <- sqldf('SELECT statehigh.name, statehigh.FES,statehigh.n,statehigh.percentage, codeoffes.FamilyTypeEmploymentStatus from statehigh,codeoffes where statehigh.FES=codeoffes.FES')
ggplot(statehigh,aes(x=name,y=percentage,fill=factor(FamilyTypeEmploymentStatus)))+xlab("The first five states which have the highest NOC")+ylab("Percentage")+geom_bar(stat="identity")+ggtitle("Percentage of FES in states with higher NOC") + theme(legend.position="bottom")
# study the last five states which have lower NOC
low1 <- select(low, FES, name) %>% na.omit()
statelow <- low1 %>% group_by(name,FES) %>% summarise(n=n()) %>% mutate(percentage = n/sum(n)*100)
statelow <- mutate(statelow, percentage = round(percentage,2))
statelow <- sqldf('SELECT statelow.name,statelow.FES,statelow.n,statelow.percentage, codeoffes.FamilyTypeEmploymentStatus from statelow, codeoffes where statelow.FES= codeoffes.FES')
ggplot(statelow, aes(x=name,y=percentage,fill=factor(FamilyTypeEmploymentStatus)))+xlab("The last five states which have the lowest NOC")+ylab("Percentage")+geom_bar(stat="identity")+ggtitle("Percentage of FES in states with lower NOC") + theme(legend.position = "bottom")
```
Compare these two graphs, we can find that States, which have higher number of children, "Both in Labor Force" and "Only husband in Labor force" have larger percentage. However, states, which have lower number of children, "Both in Labor Force" and "Female in Labor Force" have larger percentage.
Hence, we can conclude that Family Type and Employment Status will affect the number of children. The comparison in different states confirm my finding that "Only Husband in Labor Force" tend to have more children. "Female in Labor Force" tend to have less children. This finding is consistent with our assumption. Because we need money to raise babies and females take more responsibilities for raising children. 
c) Household income(Jiwen You)
```{r}

```

d) Living costs(Tian Sheng)
```{r}

```

