---
title: "EDA of children in the United States"
author: "WeiPan Cai, Yue Wang, Tian Sheng, Jiwen You, Sen Zhuang"
output:
  html_notebook: default
  html_document: default
---

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that we end up using
# in your analysis in this chunk
library(data.table)
library(ggplot2)
library(sqldf)
library(maps)
library(plyr)
library(googleVis)
library(dplyr)
library(plotly)
library(choroplethr)
library(choroplethrMaps)
```

#Summary
In this project, we have centered our exploratory data analysis around a segment of our population, families with children. We'll be investigating interesting topics such as factors that influence the number of children, states with abnormally high or low number of children and much more. Given our American community surveys on individual households, we've choosen numeric and categorical variables like Residing State, family type and employment status, languages spoken in the family and household income. Additionally, we have also created artifical new variables that extract and combind valuable information.

#Inital data overview
```{r}
# Load RData file
load(file.choose())

# Change variables to their correct class
# df$NOC <- as.factor(df$NOC) # Convert number of own children to factor
# df$NP <- as.factor(df$NP)
# df$BDSP <- as.factor(df$BDSP)
# df$RMSP <- as.factor(df$RMSP)
# df$FES <- as.factor(df$FES)
# df$HHL <- as.factor(df$HHL)
# df$WKEXREL <- as.factor(df$WKEXREL)
# df$WORKSTAT <- as.factor(df$WORKSTAT)

# print out summary stats
str(df)

```
Our dataset consists of 19 variables and total of 308611 households.

#Children across U.S. states
In the first part, we will take a look at the average number of children in households across the country.

##Map of number of children in household (here dark blue means high birth rate and white means low)
```{r}
noc_st <- df[, c('NOC', 'name')]
avg_noc_st <- aggregate(NOC~name, noc_st, FUN = mean)
avg_noc_st$name = tolower(avg_noc_st$name)
avg_noc_st$region <- avg_noc_st$name
avg_noc_st$value <- avg_noc_st$NOC
avg_noc_st <- avg_noc_st[, c(3, 4)]


# Ranked chart
sort.avg_noc_st <- avg_noc_st[order(-avg_noc_st$value),]
sort.avg_noc_st$rank <- 1:51


map_mean_noc <- state_choropleth(avg_noc_st, 
                                 title = 'Average number of children by state, 2014',
                                 legend = 'Number of own children',
                                 num_colors = 7)

map_mean_noc
```

As the plot show above, discrepences between states are obivious and there are clear clusters of states where average number of own children is praticularly high or low. For instance, new england states are themself a clusters as the number of children in those states is low. While looking at the mid-west, there is darker cluster around WY where NOW is as high as 1.9 to 2.4.

##Top 5 states with highest and lowest NOC
```{r echo=FALSE, message=FALSE, warning=FALSE}
print(sort.avg_noc_st)
```

<<<<<<< HEAD
State or district | Avg Number of children | Rank
----------------- | ---------------------  | ----
Utah              | 2.4                    | 1
Idaho             | 2.2                    | 2
Alaska            | 2.1                    | 3
North Dakota      | 2.0                    | 4
South Dakota      | 2.0                    | 5
Rhode Island      | 1.75                   | 47
Vermont           | 1.76                   | 48
West Virginia     | 1.76                   | 49
New Hampshire     | 1.72                   | 50
D.C.              | 1.66                   | 51

# Part Two
a) Languages spoken in family (Weipan Cai)
```{r}

```

b) Family Type and Employment Status (Yue Wang)

```{r, echo=FALSE}
# study the relationship between FES and NOC
working <- select(df, NOC, FES)
work1 <- working %>%
          na.omit()
size <- count(work1, FES)
# compare mean of different FES
result <- aggregate(NOC~FES, data = work1, FUN=mean)
# FamilyTypeEmploymentStatus code
fescode <- "FES,FamilyTypeEmploymentStatus
1, Both in Labor Force
2, Only Husband in Labor Force
3, Only Wife in Labor Force
4, Neither in Labor Force
5, Male in Labor Force
6, Male not in Labor Force
7, Female in Labor Force
8, Female not in Labor Force
"
codeoffes <- fread(fescode)
```

The bubble chart below shows the mean NOC under different family type and employment status. From it, we can find that Married-couple, only husband in Labor Force that has a biggest mean number of childern.

```{r}
# bubble chart
d <- diamonds[sample(nrow(diamonds),1000),]
plot_ly(d,x=result$FES, y=result$NOC, size=size$n, mode="markers") %>%
  layout(title = "Bubble chart of FES and mean of NOC", xaxis = list(title = "FES"), yaxis = list(title = "NOC"))
```

In the following part, I use ggplot to show the percentage and total NOC under different family type and employment status. From these plots, we can find that the family type that "Only Husband in Labor Force" tend to have more number of children. The family type that "Female in Labor Force" tend to have less number of children.
```{r}
work1 = tbl_df(work1)
# work2 is the table group_by(NOC,FES)
work2 <- work1 %>% group_by(NOC,FES) %>% summarise(n=n()) %>% mutate(percentage = n/sum(n) * 100)
work2 <- mutate(work2, percentage = round(percentage,2))
work2 <- sqldf('SELECT work2.NOC, work2.FES,work2.n,work2.percentage, codeoffes.FamilyTypeEmploymentStatus from work2,codeoffes where work2.FES=codeoffes.FES')
# by percentage
ggplot(work2,aes(x=NOC,y=percentage,fill=factor(FamilyTypeEmploymentStatus)))+xlab("Number of children")+ylab("Percentage")+geom_bar(stat="identity")+ggtitle("Percentage of FES under different NOC")
# by number
ggplot(work2,aes(x=NOC,y=n,fill=factor(FamilyTypeEmploymentStatus)))+xlab("Number of children")+ylab("Total Number")+geom_bar(stat="identity")+ggtitle("Total Number of FES under different NOC")
# work3 is the table group_by(FES,NOC)
work3 <- work1 %>% group_by(FES,NOC) %>% summarise(n=n()) %>% mutate(percentage = n/sum(n) * 100)
work3 <- mutate(work3,percentage = round(percentage,2))
work3 <- sqldf('SELECT work3.FES, work3.NOC,work3.n,work3.percentage, codeoffes.FamilyTypeEmploymentStatus from work3,codeoffes where work3.FES=codeoffes.FES')
# by percentage 
ggplot(work3,aes(x=FES,y=percentage,fill=factor(NOC)))+xlab("Family Type and Employment Status ")+ylab("Percentage")+geom_bar(stat="identity")+ggtitle("Percentage of NOC under different FES") + theme(legend.position="bottom")
# by number
ggplot(work3,aes(x=FES,y=n,fill=factor(NOC)))+xlab("Family Type and Employment Status ")+ylab("Total number")+geom_bar(stat="identity")+ggtitle("Total number of NOC under different FES") + theme(legend.position="bottom")
```

Then I use Sankey plot to display the relationship between Family Type Employment Status and number of children. In addition, from the ggplot above, we can find that few family will have more than four children, so our focus will be on no more than four children and married couple family type.
```{r}
# sankey plot
SKdata <- data.frame(From=c(rep("Both in Labor Force",12),rep("Only Husband in Labor Force",12),rep("Only Wife in Labor Force",12),rep("Neither in Labor Force",12),
                            rep("Male in Labor Force",12),rep("Male not in Labor Force",12),rep("Female in Labor Force",12),rep("Female not in Labor Force",12)),
                     To=c(rep(c("1","2","3","4","5","6","7","8","9","10","11","12"),8)),
                     Weight=c(40.92,41.22,13.72,3.17,0.68,0.18,0.07,0.03,0.01,0,0,0,31.47,38.59,19.52,7.11,1.99,0.77,0.29, 0.15,0.07,0.03,0.01,0,47.26,35.25,12.27,3.57,1.05,0.33,
                              0.18,0.06,0.02,0.01,0,0,49.46,30.02,12.32,5.53,1.34,0.72,0.41,0.08,0.08,0.03,0.03,0,55.22,30.64,10.24,2.93,0.74,0.16,0.06,0.01,0,0,0,0,62.77,24.31,8.81,
                              2.96,0.76,0.25,0.13,0,0,0,0,0,51.55,31.85,11.90,3.45,0.89,0.26,0.07,0.02,0.01,0,0,0,43.74,31.90,15.28,6.07,1.97,0.73,0.19,0.09,0.02,0.02,0,0.01))
Sankey <- gvisSankey(SKdata, from="From", to="To",weight="Weight",options=list(sankey="{link: {color: { fill: '#d799ae' } },
                            node: { color: { fill: '#a61d4c' },
                            label: { color: '#871b47' } }}"))
plot(Sankey)

# sankey plot about no more than four children and married couple family type
SKdata1 <-data.frame(From=c(rep("Both in Labor Force",4),rep("Only Husband in Labor Force",4),rep("Only Wife in Labor Force",4),rep("Neither in Labor Force",4)),
                     To=c(rep(c("1","2","3","4"),4)),
                     Weight=c(40.92,41.22,13.72,3.17,31.47,38.59,19.52,7.11,47.26,35.25,12.27,3.57,49.46,30.02,12.32,5.53))
Sankey1 <- gvisSankey(SKdata1, from="From", to="To",weight="Weight",options=list(sankey="{link: {color: { fill: '#d799ae' } },
                            node: { color: { fill: '#a61d4c' },
                            label: { color: '#871b47' } }}"))
plot(Sankey1)
```

Based on the first part, I try to compare the Family Type and Employment Status in states which have more children and states which have less children. 
```{r}
# Then we want to dig out more about the FES in different status.
high <- sqldf('SELECT df.NOC, df.FES, df.name from df where name in ("Utah","Idaho","Alaska","North Dakota","South Dakota")')
low <- sqldf('SELECT NOC, FES, name from df where name in ("District of Columbia","New Hampshire","West Virginia","Vermont","Rhode Island")')
# Then we want to study the relationship between state and FES
# study the fist five rank high states
high1 <- select(high, FES,name) %>% na.omit()
statehigh <- high1 %>% group_by(name,FES) %>% summarise(n=n()) %>% mutate(percentage = n/sum(n) * 100)
statehigh <- mutate(statehigh,percentage = round(percentage,2))
statehigh <- sqldf('SELECT statehigh.name, statehigh.FES,statehigh.n,statehigh.percentage, codeoffes.FamilyTypeEmploymentStatus from statehigh,codeoffes where statehigh.FES=codeoffes.FES')
ggplot(statehigh,aes(x=name,y=percentage,fill=factor(FamilyTypeEmploymentStatus)))+xlab("The first five states which have the highest NOC")+ylab("Percentage")+geom_bar(stat="identity")+ggtitle("Percentage of FES in states with higher NOC") + theme(legend.position="bottom")
# study the last five states which have lower NOC
low1 <- select(low, FES, name) %>% na.omit()
statelow <- low1 %>% group_by(name,FES) %>% summarise(n=n()) %>% mutate(percentage = n/sum(n)*100)
statelow <- mutate(statelow, percentage = round(percentage,2))
statelow <- sqldf('SELECT statelow.name,statelow.FES,statelow.n,statelow.percentage, codeoffes.FamilyTypeEmploymentStatus from statelow, codeoffes where statelow.FES= codeoffes.FES')
ggplot(statelow, aes(x=name,y=percentage,fill=factor(FamilyTypeEmploymentStatus)))+xlab("The last five states which have the lowest NOC")+ylab("Percentage")+geom_bar(stat="identity")+ggtitle("Percentage of FES in states with lower NOC") + theme(legend.position = "bottom")
```
Compare these two graphs, we can find that States, which have higher number of children, "Both in Labor Force" and "Only husband in Labor force" have larger percentage. However, states, which have lower number of children, "Both in Labor Force" and "Female in Labor Force" have larger percentage.
Hence, we can conclude that Family Type and Employment Status will affect the number of children. The comparison in different states confirm my finding that "Only Husband in Labor Force" tend to have more children. "Female in Labor Force" tend to have less children. This finding is consistent with our assumption. Because we need money to raise babies and females take more responsibilities for raising children. 

c) Household Income(Jiwen You)

In order to understand the relationship between number of children in the household with respect to the household income, firstly we plotted the histogram of the variable. Except for some extremely wealthy families, most of the households obtain annual income from 0 to $25,000. 

```{r}
hus <- na.omit(select(df,NOC,HINCP,name))
hdata <- data.frame(child=hus$NOC,income=hus$HINCP)
qplot(hdata$income,geom = "histogram",bins = 100,xlab = "Household Income", 
      main = "Histogram for Household Income",col=I("red"),alpha=I(0.3))
```

Then we generated the box plots for different NOC levels. Those wealthy households are detected as outliers in the box plot. It is interesting to see that the extremely wealthy families simply tend to have one or two children, rather than raising as many children as they can afford. In addition, it seems that the average level of household income decreases as the number of children increases. The output is much clearer when we remove those extreme values.

```{r,warning=FALSE}
plot_ly(hdata, y = income, color = factor(child), type = "box")  %>% 
  layout(title = "Boxplots for Household Income with different NOC")
```

Further more, we tried to calculate the average number of children for different household income to further verify our hypothesis. The following graph demonstrates the lowess smoothing line of the data table. There is a significant decreasing trend when the household income is greater than $50,000, while the tendency is not that clear in the remaining part. 

```{r}
# Calculate the average NOC for different income level
c <- hdata %>% group_by(income) %>% dplyr::summarise(avg_child= mean(child), 
                                                      min_child = min(child), 
                                                      max_child = max(child),
                                                      total = n())
# Smooth the average number of children for different household income
ggplot(c,aes(x=income,y=avg_child))+
  geom_smooth(aes(colour = avg_child, fill = avg_child))+
  ggtitle("Average Number of Children for Different Household Income")+
  labs(x="Household Income",y="Average Number of Children")
```

To tested our hypothesis, we used the household income from the ten states we selected previously, five with large NOC and five with small NOC. For simpleness an effectiveness of displaying the graph, we created class intervals and put our data points into their respective bins. The bin length is $1,000. And we calculated the average number of children within each bins. Different colors, grey and red, are used to distinguish the source of these household income, and the size reflects the magnitude of average number of children. Fixing the number of children, the household income in the states with less children seems to be a little be greater than that in the other five states. 

```{r,warning=FALSE}
# Select five states with large NOC and five states with small NOC to test our hypothesis
hdata_st1 <- filter(hus,name %in% c("Utah","Idaho","Alaska","North Dakota","South Dakota")) %>% mutate(code = "States with More Children")
hdata_st2 <- filter(hus,name %in% c("Rhode Island","Vermont","West Virginia","New Hampshire  ","District of Columbia")) %>% mutate(code = "States with Less Children")
hdata_st <- rbind(hdata_st2,hdata_st1) %>% mutate(HINCP=round(HINCP,-3)) %>%
  group_by(HINCP,code) %>% summarise(avg_child = mean(NOC))
plot_ly(data = hdata_st, x = HINCP, y = avg_child, mode = "markers",
        color = code, colors = "Set1", size = avg_child) %>% layout(title = "Household Income Comparision among States")
```

Except for the graphical analysis, we applied statistical test to support our result. Firstly we conducted F test to compare two variance of the household income from two group of states. The small P-value indicates the inequality of the variance. 

```{r}
var.test(hdata_st1$HINCP,hdata_st2$HINCP)
t.test(hdata_st1$HINCP,hdata_st2$HINCP)
```

Hence, we did the Welch's modified two-sample t test to test the mean. The null hypothesis is that two groups of states have same average household income. We reject the null hypothesis under 0.05 significance level. Combined with our graphical plot, we may infer that the household income and number of children are negatively correlated, even though the relationship is not that strong. 

d) Living costs(Tian Sheng)
```{r}

```

